Analysis for Biracial Grade Retention Project

### 2022-08-31 analysis/analysis.Rmd
Use counterfactual model to get at parent resource differences; add more detailed resource tables

tbl_summary %>%
  mutate_if(is.numeric, format, big.mark=",") %>%
  kbl(caption="Descriptive statistics by race. Shading indicates biracial group. Groups ordered by sample size.",
      align = c("l",rep("r", ncol(tbl_summary)-1)),
      col.names=c("race","n","% below expected grade", "mean age",
                  "% foreign-born","% speak English well",
                  "% Both parents speak english well","% either parent foreign-born",
                  "mean family income","% own home","% either parent, HS diploma",
                  "% Either parent, BA degree","% parents married")) %>%
  kable_paper() %>%
  row_spec(index_biracial, background = "#D3D3D3")
```

@@

I want a parsimonious way to visualize differences in cultural and family resources. I can sort of do it by looking at the big descriptive statistics table, but it can be a little hard to summarize. Doing each possible variable separately would be extremely tedious here, so I am going to create scores across variables for these two measures and then model them as outcomes. I can then plot the results just like I do for my actual outcome measure.

I will start by just looking at the $\alpha$ score for the correlation matrix of each case. I then create scores scaled to have a mean of zero and a standard deviation of one and fit these with a linear model using the same structure as for the models above.

## Cultural Resources

```{r cult-resource-alpha}
cult_resources <- cbind(as.numeric(!acs$foreign_born), 
                        as.numeric(!acs$foreign_born_father), 
                        as.numeric(!acs$foreign_born_mother),
                        as.numeric(factor(acs$spk_eng, 
                                          levels=c("No","Somewhat","Yes"))),
                        as.numeric(factor(acs$spk_eng_father, 
                                          levels=c("No","Somewhat","Yes"))),
                        as.numeric(factor(acs$spk_eng_mother, 
                                          levels=c("No","Somewhat","Yes"))))

alpha(cor(cult_resources))
```

That all holds together well. However, I run into a problem when I think about making a score out of this because the foreign-born status of parents actually helps to reduce the likelihood of grade retention while all of the other variables increase it. For now, I am just reverse coding it, but this leads to results that are far from intuitive. I think the simple score approach might not work well in this case.

```{r cult-resource-score}
#scale them all, add them up, and then scale again
#one catch though, I want to reverse score foreign born mother and father
#because these reduce the likelihood of grade retention, other things
#controlled
acs$cult_resources <- scale(scale(as.numeric(!acs$foreign_born))+
                              scale(as.numeric(acs$foreign_born_father))+
                              scale(as.numeric(acs$foreign_born_mother))+
                              scale(as.numeric(factor(acs$spk_eng, 
                                          levels=c("No","Somewhat","Yes"))))+
                              scale(as.numeric(factor(acs$spk_eng_father, 
                                          levels=c("No","Somewhat","Yes"))))+
                              scale(as.numeric(factor(acs$spk_eng_mother, 
                                          levels=c("No","Somewhat","Yes")))))
```

```{r cult-resource-model, warning=FALSE, fig.cap="Average cultural resources for biracial respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level. Cultural resources are determined by a summated score of standardized native-born status, parents' native-born status, English proficieny, and parents' English proficiency. Parents' native-born status is reverse coded based on model results which show it increases grade retention risk."}
#If I use the glm framework then I should be able to feed this back into
#the marg command
model_cultr <- glm(cult_resources~race+I(year-2010)*current_grade+state+metro+sex+
                    +degree_mother+degree_father+sqrt(family_income/10000)+
                     own_home+parents_married, 
                  data=acs, family="gaussian")
coef_table_cultr <- calculate_marg_means(model_cultr)

ggplot(coef_table_cultr, aes(x=term, y=estimate, color=multiracial,
                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table_cultr, multiracial))+
  geom_point()+
  coord_flip()+
  facet_wrap(~mrace, scales="free_y")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  labs(x=NULL,
       y="average adjusted cultural resource score (standard deviation units)")
```

Because of the issue with foreign-born parents, these results are not particularly intuitive. It tends to show that part-Asian and part-Latino respondents have high cultural resources, but generally not higher than the highest of their constituent groups. Other groups have results half-way in between.

## Family Resources

```{r fam-resource-score, warning=FALSE}
fam_resources <- cbind(acs$family_income, as.numeric(acs$degree_father), 
                       as.numeric(acs$degree_mother), as.numeric(acs$own_home))

alpha(cor(fam_resources))

#scale them all, add them up, and then scale again
acs$fam_resources <- scale(scale(acs$family_income)+scale(acs$own_home)+
  scale(as.numeric(acs$degree_father)+scale(as.numeric(acs$degree_mother))))
```

```{r fam-resource-model, warning=FALSE, fig.cap="Average family resources for biracial respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level. Family resources are determined by a summated score of standardized family income, parental education, and home ownership."}
#If I use the glm framework then I should be able to feed this back into
#the marg command
model_famr <- glm(fam_resources~race+I(year-2010)*current_grade+state+metro+sex+
                    foreign_born+foreign_born_mother+foreign_born_father+
                    spk_eng+spk_eng_mother+spk_eng_father, 
                  data=acs, family="gaussian")
coef_table_famr <- calculate_marg_means(model_famr)

ggplot(coef_table_famr, aes(x=term, y=estimate, color=multiracial,
                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table_famr, multiracial))+
  geom_point()+
  coord_flip()+
  facet_wrap(~mrace, scales="free_y")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  labs(x=NULL,
       y="average adjusted family resource score (standard deviation units)")
```

In most cases, family resources are roughly midway between those of the two monoracial groups. However, there are a couple of notable exceptions:

* The average family resources of Black/White students are only marginally better than those of Black students which are far below those of the average white students.
* The White/Asian group has family resources higher then either of the two constituent groups which already have the highest resources of any monoracial group. The White/Asian group has the highest family resources of any group here.
* The Indigenous/Latino and Black/Latino groups have family resources slightly lower than either of hte constituent groups, but in both cases there are hardly any differences between the monoracial groups themselves. So you would not expect controlling for these variables to change much for these groups.

>>>>

tbl_summary %>%
  mutate_if(is.numeric, format, big.mark=",") %>%
  kbl(caption="Descriptive statistics by race. Shading indicates biracial group. Groups ordered by sample size.",
      align = c("l",rep("r", ncol(tbl_summary)-1)),
      col.names=c("race","n","% below expected grade", "mean age",
                  "% foreign-born","% speak English well",
                  "% Both parents speak english well","% either parent foreign-born",
                  "mean family income","% own home","% either parent, HS diploma",
                  "% Either parent, BA degree","% parents married")) %>%
  kable_paper() %>%
  row_spec(index_biracial, background = "#D3D3D3")
```
Lets break down parent resource stuff better. First lets do SES.

```{r desc-stats-ses}
tbl_summary <- acs %>%
  group_by(race) %>%
  summarize(fam_inc=round(mean(family_income), 0),
            own_home=round(mean(own_home), 3)*100,
            par_mar=round(mean(parents_married), 3)*100,
            f_lhs=round(mean(degree_father=="LHS"), 3)*100,
            f_hs=round(mean(degree_father=="HS"), 3)*100,
            f_aa=round(mean(degree_father=="AA"), 3)*100,
            f_ba=round(mean(degree_father=="BA" | degree_father=="G"), 3)*100,
            m_lhs=round(mean(degree_mother=="LHS"), 3)*100,
            m_hs=round(mean(degree_mother=="HS"), 3)*100,
            m_aa=round(mean(degree_mother=="AA"), 3)*100,
            m_ba=round(mean(degree_mother=="BA" | degree_mother=="G"), 3)*100)

tbl_summary %>%
  mutate_if(is.numeric, format, big.mark=",") %>%
  kbl(caption="Descriptive statistics by race. Shading indicates biracial group. Groups ordered by sample size.",
      align = c("l",rep("r", ncol(tbl_summary)-1)),
      col.names=c("race","mean family income","% own home",
                  "% parents married",
                  "% father, no HS diploma","% father, HS diploma",
                  "% father, AA degree",  "% father, BA degree",
                  "% mother, no HS diploma","% mother, HS diploma", 
                  "%  mother, AA degree", "% mother, BA degree")) %>%
  kable_paper() %>%
  row_spec(index_biracial, background = "#D3D3D3")
```

```{r desc-stats-cult}
tbl_summary <- acs %>%
  group_by(race) %>%
  summarize(f_born=round(mean(foreign_born), 3)*100,
            father_fb=round(mean(foreign_born_father), 3)*100,
            mother_fb=round(mean(foreign_born_mother), 3)*100,
            spk_eng_no=round(mean(spk_eng=="No"), 3)*100,
            spk_eng_some=round(mean(spk_eng=="Somewhat"), 3)*100,
            spk_eng_yes=round(mean(spk_eng=="Yes"), 3)*100,
            f_spk_eng_no=round(mean(spk_eng_father=="No"), 3)*100,
            f_spk_eng_some=round(mean(spk_eng_father=="Somewhat"), 3)*100,
            f_spk_eng_yes=round(mean(spk_eng_father=="Yes"), 3)*100,
            m_spk_eng_no=round(mean(spk_eng_mother=="No"), 3)*100,
            m_spk_eng_some=round(mean(spk_eng_mother=="Somewhat"), 3)*100,
            m_spk_eng_yes=round(mean(spk_eng_mother=="Yes"), 3)*100)

tbl_summary %>%
  mutate_if(is.numeric, format, big.mark=",") %>%
  kbl(caption="Descriptive statistics by race. Shading indicates biracial group. Groups ordered by sample size.",
      align = c("l",rep("r", ncol(tbl_summary)-1)),
      col.names=c("race",
                  "% foreign-born","% father foreign-born",
                  "% mother foreign-born",
                  "% speak no English", "% speak some English", 
                  "% speak English",
                  "% father speak no English", "% father speak some English", 
                  "% father speak English",
                  "% mother speak no English", "% mother speak some English",
                  "% mother speak English")) %>%
  kable_paper() %>%
  row_spec(index_biracial, background = "#D3D3D3")
```

@@

A parsimonious way of doing this would be to calculate average predicted probability for each monoracial constituent group and a given biracial group, while holding constant all non-resource variables *including race*. This gives you an effective counterfactual of what the average probability would be for each group if they only differed on the resource variables. Where to set the other variables constant to might have some effect because of the transformation from logits to probs, but should not make a big difference. For convenience, I might just set it to white, but I could also try setting it to the biracial group of interest.

```{r counterfactual-ses, warning=FALSE, fig.height=12, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if there was no difference by race and the only compositional differences between groups were resources (household income, parental education, home ownership, parent and child foreign-born status, parent and child English proficiency, and whether parents are married). 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level."}
# set up the initial counterfactual dataset
counter_df <- acs %>%
  mutate(year=2014.5, # rough midpoint
         current_grade="6th", # rough midpoint from cumulative probability (55%)
         state="Minnesota", # median state effect in model 4 (Minnesota)
         metro="Metro non-central",
         sex="Male",
         race_real=race,
         race="Black")

coefs <- lapply(unique(acs$race), function(x) {
  means_margin <- marg(model4, "race", data=subset(counter_df, race_real==x))[[1]]
  return(tibble(term=x, estimate=means_margin$Margin, 
                se=means_margin$Standard.Error))
}) %>%
  bind_rows() %>%
  mutate(multiracial=str_detect(term, "/"))

coef_table <- map(as.list(str_subset(coefs$term, "/")), function(x) {
    coefs %>%
      filter(term %in% c(x, str_split(x, "/")[[1]])) %>%
      mutate(mrace=factor(x,
                          levels=c("Black/White",
                                   "White/Indigenous",
                                   "Latino/Asian",
                                   "Black/Latino",
                                   "White/Latino",
                                   "Indigenous/Latino",
                                   "Black/Asian",
                                   "White/Asian",
                                   "Indigenous/Asian",
                                   "Black/Indigenous")),
             estimate_mid=ifelse(multiracial, 
                                 mean(estimate[!multiracial]),
                                 estimate),
             se_mid=ifelse(multiracial, 
                                  sqrt(sum(se[!multiracial]^2)/4),
                                  se))
  }) %>%
    bind_rows()

  #get terms ordered
  race_names <- levels(reorder(factor(coef_table$term), coef_table$estimate))
  #pull out single race and multiple race as separate vectors
  race_single <- str_subset(race_names, "/", negate=TRUE)
  race_multi <- str_subset(race_names, "/", negate=FALSE)
  #start with single races
  race_full <- race_single
  #now loop through multiple race names and append based on the higher group
  for(race in race_multi) {
    temp <- str_split(race, "/")[[1]]
    idx <- max(which(race_full==temp[1]), which(race_full==temp[2]))
    race_full <- append(race_full, race, idx-1)
  }

  #now use this as the levels of term as a factor
  coef_table <- coef_table %>%
    mutate(term=factor(term, levels=race_full))

ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table, multiracial))+
  geom_point()+
  coord_flip()+
  facet_wrap(~mrace, scales="free_y")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="average adjusted predicted probability of being behind expected grade")

```

To summarize the results here:

|                   | Lower than both | Same as lower | Between | Same as higher | Higher than both |
|-------------------|:---------------:|:-------------:|:-------:|:--------------:|:----------------:|
| White/Asian       |       X         |               |         |                |                  |
| Black/Asian       |                 |  X            |         |                |                  |
| Black/Latino      |                 |  X            |         |                |                  |
| White/Latino      |                 |  X            |         |                |                  |
| Indigenous/Latino |                 |  X            |         |                |                  |
| Latino/Asian      |                 |  X            |         |                |                  |
| White/Indigenous  |                 |               |    X    |                |                  |
| Indigenous/Asian  |                 |               |    X    |                |                  |
| Black/Indigenous  |                 |               |    X    |                |                  |
| Black/White       |                 |               |         |         X      |                  |

In half the cases, the biracial group would have outcomes similar to the group with better resources suggesting that the parents in these matches are being positively selected. The White/Asian case is even more dramatic with a resource advantage over both single race groups. You can see this particularly for the White/Asian case in terms of family income where they have the highest of *any* group by a considerable amount. White/Asians are on average a very privileged group.

Three groups that have the "expected" in-between status based on resources all share one thing in common - they are part-Indigenous. That is not true of the Indigenous/Latino case, but is of all others. It suggests less selectivity in crossing this boundary.

Finally, the Black/White case stands out in that Black/White kids have the same risk of being held back based on their resources as do Black kids. Thus, they are not getting a resource boost nor does it seem like they are particularly positively selected.

So the effects of controlling for these resources will be complex and different across groups. In general for those that are the same as lower or lower than both, we see that it pulls their relative probability up somewhat. For the Black/White case, it pulls their probability down on the other hand. For the "between" groups its more complex and doesn't seem to do much in total.
